---
title: "Introduction to Statistics for Astronomers and Physicists"
author: "Dr Angus H Wright"
date: "Updated `r Sys.Date()`"
output:
  rmdformats::downcute:
    keep_md: false
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
    highlight: pygments
  slidy_presentation:
    toc_depth: 2
    duration: 45
    footer: '   Dr Angus H. Wright: Intro to Statistics [RUB 2021]   (Section 2c)'
    fig_width: 7
    fig_height: 5
    fig_retina: 2
    fig_caption: yes
    df_print: paged
    code_folding: hide
    incremental: true
    highlight: pygments
    theme: cosmo
  html_document:
    incremental: yes
    highlight: pygments
header-includes:
  - \usepackage{animate}
classoption: a4paper
subtitle: 'Section 2c:  Probability Distributions'
---

# Section 2: Introduction <!--{{{-->
<!--Setup {{{-->
```{r setup, include=FALSE}
rrepos <- getOption("repos")
rrepos["CRAN"] <- "https://cloud.r-project.org"
options(repos=rrepos)
options(width=100)
library(magicaxis)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = FALSE)
knitr::opts_chunk$set(class.output = "out")
knitr::opts_chunk$set(out.width="50%")
knitr::opts_chunk$set(fig.align="center")
#knitr::opts_chunk$set(fig.asp=1)
knitr::knit_engines$set(python = reticulate::eng_python)  
par(mar=c(3,3,1,1))
set.seed(666)
```
```{css, echo=FALSE}
.python { 
  background-color: 
    RColorBrewer::brewer.pal(1,"Set2");
} 
.out { 
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
```
<!--}}}-->

**Section 2: Probability & Decision Making (Weeks 3-5)**

For all aspects of modern science, an understanding of probability is required.
We cover a range of topics in probability, from decision theory and the
fundamentals of probability theory, to standard probabilistic distributions and
their origin. From this module, students will gain an insight into different
statistical distributions that govern modern observational sciences, the
interpretation of these distributions, and how one accurately models
distributions of data in an unbiased manner.

Topics include:

> + Decision theory
> + Fundamentals of probability
> + Statistical distributions and their origins

<!--}}}-->

# Probabilistic Distributions <!--{{{-->

Todays lecture is going to be all about understanding some useful probabilistic distributions. 
This list is not exhaustive, there are other standard distributions not listed here that you 
may want to investigate yourself, or come across in your own career. Nonetheless, the distributions 
that we will discuss today are all useful for understanding various problems in physics, astronomy, 
and statistics. 

## Discrete Distributions 

> + Bernoulli Random Variables
> + Geometric Distribution
> + Binomial Distribution 
> + Poisson Distribution 

## Continuous Distributions 

> + Uniform Distribution 
> + Gaussian Distribution 
> + Gamma Distribution 
> + Beta Distribution 
> + Student t-Distribution 
> + $\chi^2$-Distribution 
> + Pareto Distribution 

<!--}}}-->

# The PMF, PDF, CDF, Quantile Function <!--{{{-->

Before we begin our exploration of statistical distributions, though, we first 
need to recall some information from previous lectures, and learn two new concepts. 

## The Probability Mass and Density Functions (PMF/PDF)

We've discussed probability mass and density functions previously in the course
already, even if we haven't described them as such. The probability mass
function (or PMF) is defined as the probability of observing a discrete value
$x$ given an arbitrary sample space $\Omega$. Going back to our initial
definition of probability, this can be considered to be (in the limit of large
$n$), the relative frequency of observing the value $x$ in the set of all
outcomes.

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
set.seed(10)
magplot(density(rbinom(1E3,1,0.5),bw=0.2/sqrt(12),kern='rect',from=-0.3,to=1.3),xlab='x',ylab='PDF',
        main='PMF of a fair coin toss')
```

For non-discrete distributions, the probability of observing any particular
single value is always 0 (because there are an infinite number of possible
values in any finite interval). As such a probability mass function is
non-sensical for continuous data. Instead, the probability distribution of
continous data is defined as the relative probability of finding a value $x$
_over some finite range_. This is the **probability density function** or PDF. 

The formal definition of the PDF is therefore: 

$$
P(a\leq X \leq b) = \int_a^b p(x)\mathrm{d}x
$$
In practice we use infinitesimally small intervals to define the PDF. 

Some important properties of the PDF: 

> + The PDF is non-negative, because the $P(u)<0$ would require $\int_{u-\epsilon^{b+\epsilon} p(x)\mathrm{d}x<0$, which is impossible. 
> + The PDF must integrate to 1, which we can see if we make the interval $[a,b]=[-\infty,\infty]$: 

$$
\begin{align}
P(-\infty\leq X \leq \infty) &= \int_{-\infty}^{\infty} p(x)\mathrm{d}x \\
&= 1
\end{align}
$$

For example the PDF of train delays at a fictitious railway company that is 
totally fabricated and does not exist in Germany (or any other country) is: 
```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
set.seed(10)
magcurve(dgamma(x,shape=1.2,scale=20),from=0,to=120,,xlab='Delay (mins)',ylab='PDF',
        main='PDF of train delays at BeutscheDahn')
```
So while in reality the probability of having two delays that are _exactly_ the 
same is $\sim0$, the probability of seeing a a delay within some interval is a 
well defined function. For example, the probability of seeing a train arrive 
with no delay is $P(X=0)=0$. 

## Cumulative Distribution Function (CDF)

We've discussed the CDF somewhat at length previously. For our train delays, the 
CDF is: 
```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
set.seed(10)
magcurve(pgamma(x,shape=1.2,scale=20),from=0,to=120,,xlab='Delay (mins)',ylab='CDF',
        main='CDF of train delays at BeutscheDahn')
```
The CDF is extremely useful for measuring the fraction of probability mass that 
is being below or equal to some value $X$. In the above case, the probability 
that our train arrives with less than a $20$ minute delay is $\sim 50\%$. 

**Thanks BeutscheDahn! **

## The Quantile Function 

Again, we've looked at this somewhat at length. For analytic (invertable) functions, the 
quantile function is the inverse of the cumulative distribution function. 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
set.seed(10)
magcurve(qgamma(x,shape=1.2,scale=20),from=0.01,to=0.99,,ylab='Delay (mins)',xlab='P(X≤x)',
        main='Quantile Function of train delays at BeutscheDahn')
```
<!--}}}-->

# Expectation and Expected Values <!--{{{-->

In the first section of this course, we presented formulae for the mean,
standard deviation, variance, and covariance of a sample of observations. These
are _descriptive statistics_, in that they describe a dataset.

We can use these same statistics to describe the properties of a 
_probability distribution_ from which we draw random variables. These are
**expectations**.

The **expectation value** of a discrete variable $X$, with sample space
$\Omega$, and probability distribution $P$ is defined as:
$$
\mathbb{E}[X]=\sum_{x\in\mathcal{\Omega}} xP(X=x)
$$

For continuous data, we simply reformulate this sum as an integral: 
$$
\mathbb{E}[X]=\int_{x\in\mathcal{\Omega}} xp(x)\mathrm{d}x
$$

Let's take an example of a fair dice roll, and compute the expectation value of 
each roll. Assume a fair six-sided die ($p_i=1/6\,\,\forall\,\,i\in[1,6]$).  

$$
\begin{align}
\mathbb{E}[X]&=\sum_{x\in\mathcal{\Omega}} xP(X=x) \\
&=1\times \frac{1}{6} + 2\times \frac{1}{6} + \dots + 6\times\frac{1}{6} \\
&=\frac{1}{6}(1+2+\dots+6)\\
&=\frac{1}{n}\sum_{i=1}^{n} x_i \\
&=\bar{x}
\end{align}
$$

So the expectation value is just the arithmetic mean of the probability distribution!

We can then also calculate other statistics, like the variance:

$$
\begin{align}
\textrm{var}[X]&=\mathbb{E}[(X-\mathbb{E}[X])^2]\\
&= \mathbb{E}\left[ X^2 - 2X\mathbb{E}[X] + (\mathbb{E}[X])^2 \right]\\
&= \mathbb{E}\left[ X^2 \right] - 2\mathbb{E}[X]\mathbb{E}[X] + (\mathbb{E}[X])^2 \\
&= \mathbb{E}\left[X^2\right]-(\mathbb{E}[X])^2
\end{align}
$$
Now:
$$
\begin{align}
\textrm{var}[X]&=\mathbb{E}\left[X^2\right]-(\mathbb{E}[X])^2 \\
&=\frac{1}{n}\sum_{i=1}^{n} x^2_i - \left(\frac{1}{n}\sum_{i=1}^{n} x_i\right)^2 \\
&=
\end{align}
$$

Or the covariance: 

$$
\begin{align}
\textrm{cov}[X,Y]&=\mathbb{E}\left[(X-\mathbb{E}[X])(Y-\mathbb{E}[Y])\right] \\
&= \mathbb{E}\left[ XY + X\mathbb{E}[Y] - \mathbb{E}[X]Y + \mathbb{E}[X]\mathbb{E}[Y] \right]\\
&= \mathbb{E}\left[ XY \right] - 2\mathbb{E}[X]\mathbb{E}[Y] + \mathbb{E}[X]\mathbb{E}[Y] \\
&= \mathbb{E}\left[XY\right]-\mathbb{E}[X]\mathbb{E}[Y]
\end{align}
$$

Recall once again that, if $X$ and $Y$ are independent, $\mathbb{E}\left[XY\right]=\mathbb{E}[X]\mathbb{E}[Y]$, and so $\textrm{cov}[X,Y]=0$. 

These formulae can be used to compute the mean and variance of arbitrary probability 
distributions. We will start looking at examples of these now. 

<!--}}}-->

# The Bernoulli Random Variable <!--{{{-->

We started our discussion of probability and statistics with a simple
experiment: a single toss of a coin. When we have an experiment whose outcome is
restricted to one of two outcomes ("success/failure", "yes/no") the variable $X$
is said to be a Bernoulli random variable.

The Bernoulli random variable is defined as taking the value $1$ with a
probability $p$, and the value $0$ with a probability $1-p$. Therefore the PMF
of a Bernoulli Random Variable is:
$$
X\sim \textrm{Bern}(p) \\
P(X=x)=
\begin{cases}
1-p & \textrm{if } x=0\\
p & \textrm{if } x=1\\
\end{cases}
$$
The related cumulative distribution function is therefore: 
$$
F(x)=
\begin{cases}
0 & \textrm{if } x<0 \\
1-p & \textrm{if } 0\leq x<1 \\
1 & \textrm{if } x\geq 1
\end{cases}
$$

The Bernoulli Random Variable has the following useful properties: 

> + Expectation: $\mathbb{E}[X]=p$

$$
\begin{align}
\mathbb{E}[X] &= \sum_{x\in\mathcal{\Omega}} xP(X=x) \\
&=  0\times(1-p) + 1\times p \\
&= p
\end{align}
$$

> + Variance: $\textrm{var}[X]=p(1-p)$. 

$$
\begin{align}
\textrm{var}[X] &= \mathbb{E}\left[X^2\right]-(\mathbb{E}[X])^2 \\
&= \sum_{x\in\mathcal{\Omega}} \left[x^2P(X=x^2)\right] - \left[\sum_{x\in\mathcal{\Omega}} xP(X=x)\right]^2  \\
&=  \left[0^2\times(1-p) + 1^2\times p\right] - p^2 \\
&= p-p^2 \\
&= p(1-p)
\end{align}
$$

If $p=0$ or $p=1$, then the Bernoulli random variable follows the **degenerate distribution**; a distribution with only one outcome (and so no randomness is possible). The degenerate distribution can 
be trivially seen from the above to have mean $p$ and variance $0$, which makes sense! 

<!--}}}-->

# The Geometric Distribution <!--{{{-->

We have a biased coin, that throws heads with a probability $P(\{H\})=p$. We flip this coin _until_ the first head appears. The number of flips that we have to make is a discrete random variable that takes integer values greater than or equal to $1$. 

Each toss of the coin has the probability $p$ of being a head, and so has a probability $(1-p)$ of being a tail. In order for us to observe $n$ tosses, we must have thrown $(n-1)$ tails in a row, and then a head. Therefore the probability of seeing $n$ tosses is: 

$$
P(X=n)=(1-p)^{(n-1)}p.
$$
This is the PMF of the **geometric distribution**. It is defined only for positive integers $n\geq 1$ and $0\leq p \leq 1$. 

The geometric distribution describes probability of observing a number of independent trials, which have some probability of success $p$, prior to observing a success (at which point your experiment ends). 

> + Expectation: $\mathbb{E}[X]=p$
> + Variance: $\textrm{var}[X]=p(1-p)$. 

<!--}}}-->

# The Binomial Distribution <!--{{{-->

A different statistic that may be of interest is how often you expect your biased coin to throw heads given $n$ tosses. 
In this case we're asking how many times $k$ do I observe 'success' in $n$ trials. The probability of 
seeing $k$ successes from $n$ trials: $p^k$. Consequently the probability of seeing the $n-k$ failues is: $(1-p)^{n-k}$. Finally, there are: 
$$
\frac{n\times (n-1)\times(n-2)\times\dots\times(n-k+1)}
{k\times (k-1)\times(k-2)\times\dots\times1}
$$  
ways of constructing $k$ successes from $n$ trials (when order of appearance doesn't matter). 
Therefore, the probability of observing $k$ successes from $n$ trials with our coin is: 

$$
k\sim \textrm{Bin}(n,p) \\ 
P(k; n, p) = \frac{n!}{k!(n-k)!}\,p^k(1-p)^{n-k},
$$
This is the PMF of the **binomial function**. 

The observant here will recognise the coefficient from last lecture. This is called the *binomial coefficient*, which we presented as being read as `$n$ choose $k$':  

$$
P(k; n, p) = {n\choose k}p^k(1-p)^{n-k}.
$$

The binomial function has the properties: 

> + Expectation: $$\mathbb{E}[X]=Np$
> + Variance: $\textrm{var}[X]=Np(1-p)$
 
## The Multinomial Distribution  <!--{{{-->

The natural extension of the binomial distribution is to allow for multiple binomial 
outcomes, like in our urn model. 

If we perform $N$ independent realisations of an experiment with $k\in\Omega$ outcomes, where the 
$i^\textrm{th}$ outcome has probability $p_i$: the probability of observing outcome $1$ $n_1$ times, outcome $2$ $n_2$ times, etc, follows the multinomial distribution PMF: 

$$
n_1,n_2,\dots,n_k \sim \textrm{Multi}(N,p_1,p_2,\dots,p_k)\\
P(n_1, n_2, \dots,n_k; N, p_1,p_2,\dots,p_k) = \frac{N!}{n_1!n_2!\dots n_k!}\,p_1^{n_1}p_2^{n_2}\dots p_k^{n_k},
$$ 
for non-negative integers $n_1,n_2,\dots,n_k$ and where $\sum_{i=1}^k n_i = N$. 

We can use the multinomial distribution, for example, so get the PMF of our 
2-dice roll that we studied in previous lectures: 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magplot(density(colSums((1:6)*rmultinom(n=1e5, size=2, prob=rep(1/6,6))),from=1,
                to=13,bw=0.2/sqrt(12),kern='rect'),xlab="Sum of two dice", 
        ylab='PDF',main='Multinomial PMF of the Sum of Two Dice')
```

The multinomial distribution has the properties: 

> + Expectation: $\mathbb{E}[X_i]=Np_i$
> + Variance: $\textrm{var}[X_i]=Np_i(1-p_i)$

<!--}}}-->

<!--}}}-->

# The Poisson Distribution <!--{{{-->

Consider an experiment where we want to know the number of occurrences of an event 
within a fixed time window: say, the number of atomic decay processes in a box per 
hour. Because we are interested in counts per hour, the value of our outcomes are 
non-negative integers. 

We can make two additional constraints: 

> + The decay processes happen with a fixed rate; and
> + The time between decay processes is independent of the interval between previous decays.

Given this experiment, the resulting observations will follow the **Poisson distribution**, 
which has the PMF: 
$$
k\sim \textrm{Po}(\lambda) \\
P(k; \lambda) = e^{-\lambda}\frac{\lambda^k}{k!},
$$
where $\lambda$ is the so-called `intensity' parameter which governs the shape of the distribution. We can calculate 
the expectation of the Poisson distribution using our standard formula (but this is 
a bit more involved than previously...): 

$$
\begin{align}
\mathbb{E}[X] &= \sum_{x\in\mathcal{\Omega}} xP(X=x) \\
&=  \sum_{x=1}^{\infty} x\frac{e^{-\lambda}\lambda^x}{x!} \\
&=  \sum_{x=1}^{\infty} \frac{e^{-\lambda}\lambda^x}{(x-1)!} \\
&=  \lambda e^{-\lambda}\sum_{x=1}^{\infty} \frac{\lambda^{(x-1)}}{(x-1)!} \\
&=  \lambda e^{-\lambda}\left(\frac{\lambda^0}{0!}+ \frac{\lambda^1}{1!} + \dots \right)\\
&=  \lambda e^{-\lambda}\sum_{x=0}^{\infty} \frac{\lambda^x}{x!} \\
&=  \lambda e^{-\lambda}e^{\lambda} \\
&=  \lambda
\end{align}
$$
So the $\lambda$ parameter describes the **expectation** of the Poisson distribution!

We can perform the same exercise to calculate the variance (you can try this yourself), 
and we find that the **variance** of the Poisson distribution is _also_ $\lambda$. 

The Poisson distribution has the property: 

> + $\mathbb{E}[X]=\textrm{var}[X]=\lambda$. 

The PMF of a Poissonion random variable, for a range of values of $\lambda$ is: 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magplot(0:25,dpois(0:25,10),type='h',col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,0.4),xlab='x',ylab='PDF')
lines(0:25,dpois(0:25,5),type='h',col=seqinr::col2alpha('green3',0.5),lwd=4)
lines(0:25,dpois(0:25,1),type='h',col=seqinr::col2alpha('blue2',0.5),lwd=4)
legend('topright',legend=c(expression(paste("X~Po(",lambda,"=10)")),expression(paste("X~Po(",lambda,"=5)")),expression(paste("X~Po(",lambda,"=1)"))),lty=1,col=c('red2','green3','blue2'),lwd=2,inset=0.1,bty='n')
```

The Poisson distribution is very important in the natural sciences, as signals 
that we measure are frequently actually integrated counts over a period of time. 
In astronomy, for example, observed fluxes of galaxies are determined by the number of 
photons received at the detector during an observation. Therefore the minimal uncertainty 
is determined by the Poisson distribution. This is particularly relevant in the 
high-energy regime, because $\gamma$-ray photons are relatively rare. As such, 
flux uncertainties at these short-wavelengths are generally Poisson-noise dominated. 

One further important point about the Poisson distribution we can determine using 
some intuition. 

Consider a Poisson variable. The expected number of occurances of an event in some 
period $t$ is $\lambda$. How many events would you intuitively expect to observe in a 
period $2t$? 

+ $2\lambda$ is the sensible result, given both intuition and what we know about the 
expectation value. 

But the poisson variable need not be constrained to _temporal_ events; I can just as 
easily construct a poisson random variable that determines the number of events 
observed in a spatial interval $x$. Take, e.g., the number of insects per unit 
area in a field. If the insect movements are random and independent, then the 
number of insects per unit area ought to be a random poisson variable. 

But this means that in twice the area we ought to find twice the number of
insects. Putting this into more rigorous mathematical context: In some
multi-dimensional domain $\mathcal{D}$, the number of spatially random
observations within a subset $s$ of $\mathcal{D}$ is a Poisson random variable
with intensity $m(s)\lambda$, where $m(s)$ is the area/volume of $s$.

Models such as these are useful in Physics and Astronomy because they capture 
the properties that: 

> + the within the domain are randomly distributed; and
> + the probability that you find a point doesn’t depend on where you are in the 
domain. 

<!--}}}-->

# Continuous Distributions 

Until now we have looked exclusively at discrete distributions. We will now shift 
focus to continuous distributions. 

# The Uniform Distribution <!--{{{-->

The first continuous distribution that we will examine is the most simple. 
The **uniform distribution** is characterised by a lower bound $a$ and upper bound $b$ 
(where $a<b$), with constant probability density in between: 

$$
X\sim U(a,b) \\
p(X=x)=
\begin{cases}
0 & \textrm{if } x<a \\
\frac{1}{b-a} & \textrm{if } a\leq x\leq b \\
0 & \textrm{if } x>b
\end{cases}
$$

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dunif(x,0,1),from=-6,to=6,col=seqinr::col2alpha('red2',0.5),lwd=4,type='s',
        ylim=c(0,1.1),xlab='x',ylab='PDF',main="Uniform Distribution with various limits")
curve(dunif(x,-2,1.5),from=-6,to=6,col=seqinr::col2alpha('green3',0.5),lwd=4,add=T,type='s')
curve(dunif(x,-5,5),from=-6,to=6,col=seqinr::col2alpha('blue2',0.5),lwd=4,add=T,type='s')
legend('topleft',legend=c(expression(paste("X~U(a=0,b=1)")),expression(paste("X~U(a=-2,b=1.5)")),
                           expression(paste("X~U(a=-5,b=5)"))),lty=1,
       col=c('red2','green3','blue2'),lwd=2,inset=0.05,bty='n')
```

The Uniform distribution has properties: 

> + Expectation $\mathbb{E}[X]=\frac{a+b}{2}$
> + Variance $\textrm{var}[X]=\frac{(b-a)^2}{12}$

The uniform distribution is useful when considering random events over a fixed interval. 
For example: Consider the scenario where the tram to RUB arrives at Bochum
Hauptbahnhof every 5 minutes during peak travel times. If people arrive at the station 
randomly and without knowing the schedule, how long can they expect to wait for a tram? 

The waiting times in this scenario is a random uniform variable: 

$$
X\sim U(0,5) \\
p(X=x)=
\begin{cases}
\frac{1}{5} & \textrm{if } 0\leq x\leq 5 \\
0 & \textrm{otherwise}
\end{cases}
$$
The expectation of this distribution is $\mathbb{E}[X]=\frac{5}{2}$, so the 
expected waiting time is $2.5$ minutes. Similarly, we can compute the probability 
of waiting less than $1$ minute using the cumulative probability distribution 
(which for the uniform distribution is trivial): 

$$
P(X<x)=
\begin{cases}
0 & \textrm{if } x<a \\
\frac{x-a}{b-a} & \textrm{if } a\leq x\leq b \\
1 & \textrm{if } x>b
\end{cases}
$$
So for our train waiting times: 
$$
P(X<x)=
\begin{cases}
0 & \textrm{if } x<a \\
\frac{x}{5} & \textrm{if } 0\leq x\leq 5 \\
1 & \textrm{if } x>5
\end{cases}
$$
which means that there is a $20\%$ chance of waiting less than $1$ minute. 

<!--}}}-->

# The Beta Distribution <!--{{{-->

The second continuous distribution that we're going to look at is the **beta distribution**. 
The beta distribution is not one that you are likely familiar with, but it is useful for 
two reasons: 

> + It is an extremely flexible function that is defined over a finite range; and 
> + It is widely used in congugate analyses (which we will explore in the next section of the course). 

The Beta distribution is defined as: 
$$
X\sim\textrm{Be}(a,b)\\
p(X=x;a,b)=\frac{\Gamma(a) \Gamma(b)}{\Gamma(a+b)}x^{a-1}(1-x)^{b-1},
$$
where $\Gamma$ is the Gamma function, defined as: 
$$
\Gamma(n)=(n-1)!
$$
for integer values $n$, and: 
$$
\Gamma(z)=\int_0^\infty x^{z-1}e^{-x}\mathrm{d}x
$$
for any complex number with a positive real component $\mathcal{R}(z)>0$. 

The variability of the Beta distribution as a function of shape parameters $a$ and $b$ are shown below: 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dbeta(x,1,1),from=0,to=1,col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,10),xlab='x',ylab='PDF',main="Beta Distribution with identical shape params")
curve(dbeta(x,10,10),from=0,to=1,col=seqinr::col2alpha('green3',0.5),lwd=4,add=T)
curve(dbeta(x,50,50),from=0,to=1,col=seqinr::col2alpha('blue2',0.5),lwd=4,add=T)
legend('topright',legend=c(expression(paste("X~Be(a=1,b=1)")),expression(paste("X~Be(a=10,b=10)")),
                           expression(paste("X~Be(a=50,b=50)"))),lty=1,
       col=c('red2','green3','blue2'),lwd=2,inset=0.1,bty='n')
```
```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dbeta(x,1,10),from=0,to=1,col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,12),xlab='x',ylab='PDF',main="Beta Distribution with different shape params")
curve(dbeta(x,10,1),from=0,to=1,col=seqinr::col2alpha('green3',0.5),lwd=4,add=T)
curve(dbeta(x,3,15),from=0,to=1,col=seqinr::col2alpha('blue2',0.5),lwd=4,add=T)
curve(dbeta(x,20,100),from=0,to=1,col=seqinr::col2alpha('orange2',0.5),lwd=4,add=T)
legend('topright',legend=c(expression(paste("X~Be(a=1,b=10)")),expression(paste("X~Be(a=10,b=1)")),
                           expression(paste("X~Be(a=3,b=15)")),expression(paste("X~Be(a=20,b=100)"))),lty=1,
       col=c('red2','green3','blue2','orange2'),lwd=2,inset=0.1,bty='n')
```

The Beta distribution has the properties: 

> + Expectation $\mathbb{E}[X]=\frac{a}{a+b}$
> + Variance $\textrm{var}[X]=\frac{ab}{(a+b)^2(a+b+1)}$

<!--}}}-->

# The Gamma Distribution <!--{{{-->

Similar to the Beta distribution, the **gamma distribution** will become relevant 
later in the course due to its use in conjugate analyses. The gamma distribution 
is defined by two parameters, the shape parameter $a$ and the scale parameter $s$.
The gamma distribution PDF is defined as: 
$$
X\sim \textrm{Ga}(a,s) \\
p(X=x; a, s) = \frac{s^a x^{a -1}e^{-\beta x}}{\Gamma(\alpha)}. 
$$
As with the Beta distribution, the Gamma distribution is extremely flexible. In 
fact, the Gamma distribution encompasses (as special cases with particular parameter 
combinations) many interesting distributions that you will likely already be
familiar with, such as the exponential distribution and the $\chi^2$-distribution 
(the latter of which we will discuss in a few slides time!). 

The variability of the Gamma distribution as a function of shape parameter $a$ and scale 
parameter $s$ are shown below: 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dgamma(x,1,1),from=0,to=10,col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,2),xlab='x',ylab='PDF',
        main="Gamma Distribution with identical shape and scale params")
curve(dgamma(x,5,5),from=0,to=10,col=seqinr::col2alpha('green3',0.5),lwd=4,add=T)
curve(dgamma(x,15,15),from=0,to=10,col=seqinr::col2alpha('blue2',0.5),lwd=4,add=T)
legend('topright',legend=c(expression(paste("X~Ga(a=1,s=1)")),expression(paste("X~Ga(a=5,s=5)")),
                           expression(paste("X~Be(a=15,s=15)"))),lty=1,
       col=c('red2','green3','blue2'),lwd=2,inset=0.1,bty='n')
```

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dgamma(x,1,5),from=0,to=10,col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,2),xlab='x',ylab='PDF',main="Gamma Distribution with different shape and scale params")
curve(dgamma(x,5,1),from=0,to=10,col=seqinr::col2alpha('green3',0.5),lwd=4,add=T)
curve(dgamma(x,15,3),from=0,to=10,col=seqinr::col2alpha('blue2',0.5),lwd=4,add=T)
curve(dgamma(x,20,10),from=0,to=10,col=seqinr::col2alpha('orange2',0.5),lwd=4,add=T)
legend('topright',legend=c(expression(paste("X~Ga(a=1,s=5)")),expression(paste("X~Ga(a=5,s=1)")),
                           expression(paste("X~Ga(a=15,s=3)")),expression(paste("X~Ga(a=20,s=10)"))),lty=1,
       col=c('red2','green3','blue2','orange2'),lwd=2,inset=0.1,bty='n')
```

The Gamma Distribution has properties: 

> + Expectation $\mathbb{E}[X]=\frac{a}{s}$
> + Variance $\textrm{var}[X]=\frac{a}{s^2}$

<!--}}}-->

# The Gaussian Distribution <!--{{{-->

Everyone in this course is undoubtably familiar with the **Gaussian Distribution**,
and it is one of the most important distributions in statistics. 

A random variable $X$ which follows a Gaussian distribution with mean $\mu$ and variance 
$\sigma^2$ has the PDF: 
$$
X\sim N(\mu,\sigma^2) \\
p(X=x; \mu,\sigma^2) = \frac{1}{\sqrt{2 \pi \sigma^2}}e^{-\frac{(x - \mu)^2}{2 \sigma^2}},
$$

The cumulative distribution function of the Gaussian distribution is widely used 
because of the frequency with which variables are modelled as being Gaussian. The 
CDF is defined as: 
$$
\Phi(X=x)=\int_{-\infty}^{x} p(t)\mathrm{d}t
$$
There is no analytic formulae for the CDF of the $N(\mu,\sigma^2)$ gaussian, and 
so the values are computed using numerical and computational methods. 

A useful special case of the Gaussian distribution is the so-called "standard" Gaussian: 
$X\sim N(0,1)$:
$$
X\sim N(0,1) \\
p(X=x) = \frac{1}{\sqrt{2 \pi}}e^{-\frac{x^2}{2}},
$$

It is possible to transform any gaussian variable into a standard gaussian variable using 
the so-called $Z$-transformation: 
$$
X\sim N(\mu,\sigma^2) \\
Z=\frac{X-\mu}{\sigma} \\ 
\therefore Z\sim N(0,1)
$$

This is useful for many reasons, particularly in modelling and understanding biases, 
but here we will use the transformation to derive some useful properties of the gaussian 
distribution: 

$$
\begin{align}
P(X\leq b) &= P\left(\frac{X-\mu}{\sigma} \leq \frac{b-\mu}{\sigma}\right)  \\
&= P\left(Z \leq \frac{b-\mu}{\sigma}\right) \\
&= \Phi(\frac{b-\mu}{\sigma}).
\end{align}
$$

> + The probability of a Gaussian random variable being less than some value $b$, for an 
arbitrary mean and variance, is equal to the probability mass below the equivalent Z-score in the standard gaussian. 

By symmetry and integral constraints we can equally demonstrate that: 

$$
P(X > a) = 1 - \Phi(\frac{a-\mu}{\sigma}); \\
P(a \leq X \leq b) = \Phi(\frac{b-\mu}{\sigma}) - \Phi(\frac{a-\mu}{\sigma}); \\
\Phi(-a) = 1-\Phi(a); \\
P(-a < Z < a) = 2\Phi(a)-1. 
$$

The Gaussian Distribution has properties: 

> + Expectation $\mathbb{E}[X]=\mu$
> + Variance $\textrm{var}[X]=\sigma$

<!--}}}-->

## The Multivariate Gaussian Distribution  <!--{{{-->

One can extend the Gaussian Distribution to multiple dimensions, forming the **multivariate gaussian distribution**. The multivariate Gaussian distribution is characterised 
by a vector of means $\mu$ and a **covariance matrix** $\Sigma$: 
$$
X \sim N_k(\mu, \Sigma),
$$

The primary difference between the description of the 
univariate and multivariate Gaussian distributions is therefore that the
multivariate gaussian requires the specification of the covariance matrix, which
we discussed in previous lectures:
$$
\Sigma = 
\begin{pmatrix}
{\rm var}(X_1)  & {\rm cov}(X_1,X_2) & \dots & {\rm cov}(X_1,X_n) \\
{\rm cov}(X_2,X_1)  & {\rm var}(X_2) & \dots & {\rm cov}(X_2,X_n) \\
\vdots & \vdots & \ddots & \vdots \\
{\rm cov}(X_n,X_1)  & {\rm cov}(X_n,X_2) & \dots & {\rm var}(X_n)
\end{pmatrix}
$$
The covariance matrix encodes the degree of mutual variance between the different 
gaussian components. In cases with zero covariance: 

$$
X\sim N_2(\mu,\Sigma)\\
\mu=\{-2,3\} \\
\Sigma = 
\begin{pmatrix}
4  & 0 \\
0  & 1 \\
\end{pmatrix}
$$

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
library(mvtnorm)
#Zero covariance
cov=cbind(c(4,0),c(0,1)) 
magplot(rmvnorm(1e3, mean=c(-2,3), sigma=cov),xlab=expression(X[1]),ylab=expression(X[2]),asp=1)
``` 

Or the case of significant covariance: 
$$
X\sim N_2(\mu,\Sigma)\\
\mu=\{-2,3\} \\
\Sigma = 
\begin{pmatrix}
4  & 2 \\
2  & 1 \\
\end{pmatrix}
$$

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
#Non-zero covariance
cov=cbind(c(4,1.1),c(1.1,1)) 
magplot(rmvnorm(1e3, mean=c(-2,3), sigma=cov),xlab=expression(X[1]),ylab=expression(X[2]),asp=1)
``` 

<!--}}}-->

# What's with all the Gaussians?! <!--{{{-->

In the natural sciences, we see the Gaussian distribution _a lot_. We typically formulate our uncertainties as Gaussian, 
we model physical processes as Gaussian, and we quote significances in terms of Gaussians. **Why do we do this??**

## Central Limit Theorem (CLT) <!--{{{-->

In a simplified manner, the central limit theorem states that, for an arbitrary distribution with mean $\mu$ and 
variance $\sigma^2$, we can draw $n$ observations at random and calculate their mean $M_n$. The distribution 
$$
\left[\frac{M_n-\mu}{\sigma/\sqrt{n}}\right]\rightarrow N(0,1)
$$
as $n\rightarrow\infty$. 

This is a pretty remarkable statement. It means that (in certain conditions that I don't raise here, but 
just know that they are met in _many_ circumstances), averaging over random variables will produce a Gaussian distribution 
of results _regardless of the shape of the distribution the samples are drawn from_!

Let's do an experiment. In our experiment, we observe a random variable that follows a Gamma distribution: 
$$ 
X\sim Ga(1,2)
$$
```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dgamma(x,1,2),from=-2,to=10,col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,2),xlab='x',ylab='PDF',main="X~Ga(1,2)")
```

It should go without saying: the gamma distribution we have picked doesn't look very gaussian. But what happens if we are interested in the properties of sub-samples of our population, say, the mean of sets of $2$ observations? Or $4$? Or $16$?

The expectation of this distribution is $\mathbb{E}[X]=\frac{a}{s}=0.5$ (from earlier). If we take our observations and compute the expectation of discrete subsets of the observations we find the below: 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dgamma(x,1,2),from=-1,to=3,col=seqinr::col2alpha('red2',0.5),lwd=4,
        ylim=c(0,3.5),xlab='x',ylab='PDF',main="X~Ga(1,2)")
mat<-matrix(rgamma(200*2,1,2),nrow=200)
lines(density(rowMeans(mat),from=-1,to=3,bw=0.2/sqrt(12),kern='rect'),col=seqinr::col2alpha('green3',0.5),lwd=4)
mat<-matrix(rgamma(200*4,1,2),nrow=200)
lines(density(rowMeans(mat),from=-1,to=3,bw=0.2/sqrt(12),kern='rect'),col=seqinr::col2alpha('blue2',0.5),lwd=4)
mat<-matrix(rgamma(200*16,1,2),nrow=200)
lines(density(rowMeans(mat),from=-1,to=3,bw=0.2/sqrt(12),kern='rect'),col=seqinr::col2alpha('orange2',0.5),lwd=4)
legend('topright',legend=c("Initial Distribution",expression(M[2]),expression(M[4]),
                           expression(M[16])),lty=1,
       col=c('red2','green3','blue2','orange2'),lwd=2,inset=0.1,bty='n')
```

To demonstrate just _how good_ the approximation is, we can standardise the distributions and overplot a $N(0,1)$ gaussian: 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
magcurve(dnorm(x,0,1),from=-4,to=4,col=seqinr::col2alpha('black',0.5),lwd=4,
        ylim=c(0,0.7),xlab='x',ylab='PDF')
lines(density(scale(rowMeans(mat)),from=-4,to=4,bw=1/sqrt(12),kern='rect'),col=seqinr::col2alpha('orange2',0.5),lwd=4)
legend('topright',legend=c("X~N(0,1)",expression(frac(M[16]-mu,sigma/sqrt(n)))),lty=1,
       col=c('black','orange2'),lwd=2,inset=0.1,bty='n')
```

This result has profound implications for the experimental sciences. It is the 
reason why we model complex systems as Gaussian distributions. It gives rise 
to Gaussian distributions in systems where repeat random events are the norm 
(e.g. repeated scattering of particles drives their energy to the expectation), 
causes us to model our uncertainties as gaussians (because final observed quantities 
are the cumulative result of many underlying processes), and leads us to 
compare models by quoting "sigma" differences. 

Finally, one last comment about the CLT. The tails of the gaussian distribution are 
remarkable tight: we expect more than $95\%$ of gaussian observations to reside 
within $2$ standard deviations of the mean. However in the CLT, the convergence 
happens most rapidly in the _centre_ of the distribution. This means that, in the 
limit of small numbers of observations, we expect the CLT converge on the Gaussian distribution 
fastest at the centre of the distribution, and slowest in the wings. 

> + As a result, gaussian modelling distributions of objects which only enact few averagings will 
tend to ignore the presence of the outlier that are expected, leading to problems. **Always pay attention 
to the tails of your distributions**. 

<!--}}}-->

<!--}}}-->

# The Student t-Distribution <!--{{{-->

Given our observation about the wings of the convergence to a gaussian in the CLT, 
we are motivated to explore different distributions that account for this behaviour. 

We can define a new statistic $t$ which is our mean statistic from before: 
$$
t = \frac{\bar{x}-\mu}{s/\sqrt{N}},
$$

Given the $t$ variable, we can compute the **student t-distribution**: 
$$
p(t; N)	=	\frac{\Gamma(\frac{1}{2} N)}{\sqrt{\pi (N-1)} \Gamma\left[\frac{1}{2} (N-1)\right]}\left(1+\frac{t^2}{N-1}\right)^{-N/2}
$$
Frequently, though, we will see this distribution parameterised in terms of the 
'degrees of freedom' $\nu=(N-1)$.  

Nonetheless, this still looks complicated, but the limits are the most relevant
for understanding. When $N$ is large, the student $t$-distribution converges to
the gaussian distribution, as we wanted. However, for small $N$, the
distribution converges to the Lorentzian (or Cauchy) distribution, which
contains the strong wings that we expect.

We can demonstrate the above limiting behaviour easiest when computing the
t-statistic for initially gaussian observations (rather than our one-sided Gamma
distribution from last slide). We perform exactly the same game, though. We
average $n$ observations from an initially Gaussian distribution, and plot the
distribution of our $t$ statistic above (with their centroids shifted for
clarity):
```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
library(matrixStats)
nobs<-1e4
n<-c(3,5,10,100)
cols<-c("blue3","green3","red3","orange3")
legend<-c("Gaussian","Student t")
magplot(1,type='n', col='blue2',lwd=2, xlim=c(-4,5+length(n)), ylim=c(0,0.45), 
        xlab='t', ylab='PDF')
for (i in 1:length(n)) { 
  mat<-matrix(rnorm(n[i]*nobs),ncol=n[i])
  stat<-rowMeans(mat)/(rowSds(mat)/sqrt(n[i]))
  lines(density(stat+i,bw=0.2,from=-4,to=6+length(n)),col=cols[i],lwd=2)
  curve(dt(x-i, df=n[i]-1), from=-4, to=6+length(n), col=cols[i],lwd=2, lty=2, add=TRUE)
  curve(dnorm(x-i), from=-4, to=6+length(n), col=cols[i],lwd=2, lty=3, add=TRUE)
  legend<-c(legend,bquote(M[.(n[i])]+.(i)))
}
legend('topright',legend=as.expression(legend),
       col=c('black','black',cols),lty=c(3,2,rep(1,length(n))),lwd=2,inset=0.05,bty='n')
```

The Student t-Distribution has properties: 

> + Expectation $\mathbb{E}[X]=\frac{a}{s}$
> + Variance $\textrm{var}[X]=\frac{a}{s^2}$

<!--}}}-->

# The $\chi^2$ Distribution <!--{{{-->

Again related to the CLT is the **$\chi^2$ distribution**. The $\chi^2$
distribution is that which we find when summing random _squared deviations_ from a
Normal distribution. The _$\chi^2$ distribution_ is extremely common in the experimental 
sciences, as it is the classic metric that is used in the quantification of 
_goodness-of-fit_ of a model to data. In this application it is frequently 
referred to as the Pearson $\chi^2$ statistic. 

The $\chi^2$ statistic is defined as:
$$
\chi^2 \equiv C = \sum\limits_{i=1}^N \frac{(x_i - \mu)^2}{\sigma^2}.
$$
For the $\chi^2$ distribution the degrees of freedom is given by $\nu=N$. 
The $\chi^2$ distribution is then defined as:

$$
p(C; \nu) = \frac{1}{2^{\nu/2} \Gamma(\nu/2)} C^{\nu/2-1} e^{-C/2}
$$
 
Use of the $\chi^2$ statistic in quantifying goodness-of-fit is frequently summarised 
in the literature as _reduced $\chi^2$_, generally formulated as $\chi^2_{\rm r}=\chi^2/\nu_{}$. 
The choice of this statistic is that the expectation of the $\chi^2$ variable is 
$\nu$: $\mathbb{E}[\chi^2/\nu]= 1$. 
However, model fits in the literature generally perform maximum likelihood optimisation, 
which aims to find the _mode_ of the PDF, rather than to produce samples that 
report the mean/expectation. 
Looking at the 
$\chi^2$ distribution, it's clear that the distribution is asymmetric, and so these 
results will converge on different answers: 
```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE}
library(matrixStats)
magplot(1,type='n', col='blue2',lwd=2, xlim=c(0,85), ylim=c(0,0.25), 
        xlab=expression(chi^2), ylab='PDF')
n<-c(3,10,30,60)
legend<-"Mean"
for (i in 1:length(n)) { 
  curve(dchisq(x, df=n[i]), from=0, to=85, col=cols[i],lwd=2, lty=1, add=TRUE)
  abline(v=n[i],lty=2,col=cols[i])
  legend<-c(legend,bquote(chi^2*(.(n[i]))))
}
legend('topright',legend=as.expression(legend),
       col=c('black',cols),lty=c(2,rep(1,length(n))),lwd=2,inset=0.05,bty='n')
```

The **mode** of the distribution is actualy at $\nu-2$. Therefore, the appropriate formulation of the 
reduced $\chi^2$ is for maximum likelihood fits is: 
$$
\chi^2_{\rm r}=\frac{\chi^2}{\nu-2}_{}
$$

The $\chi^2$ Distribution has properties: 

> + Expectation $\mathbb{E}[X]=\nu$
> + Variance $\textrm{var}[X]=2\nu$

 
<!--}}}-->

# Wealth and equality <!--{{{-->

Consider $1,000$ people who each have $\$1,000$. We take two of these people at 
random, and give a fixed fraction of their total worth to the other (think of 
this a transaction of some kind). We then repeat this process $100,000$ times. 

What will the final distribution of wealth look like? 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE,warnings=FALSE}
curve(dnorm(x,1e3,100),from=0,to=2e3,lwd=2, type='s',
        xlab='Net Worth per person', ylab='PDF',col='blue3')
curve(dunif(x,300,1700),from=0,to=1e4,lwd=2,add=TRUE,type='s',
        xlab='Net Worth per person', ylab='PDF',col='red3')
curve(dexp(x,1/1e3),from=0,to=1e4,lwd=2,add=TRUE,type='s',
        xlab='Net Worth per person', ylab='PDF',col='green3')
legend('topright',lty=1,lwd=2,col=c('blue3','red3','green3'),
       legend=c("Option A","Option B","Option C"),inset=0.05)

```

<!--}}}-->

# Can you get away with tax fraud? <!--{{{-->

For totally legitimate, not-illegal reasons: you need to fabricate a series of 
transactions to "pad out" a table of real transactions, such that your fabricated 
numbers are indistinguishable from "real" transactions. The transaction
values range from large (many thousands of dollars) to small (a few cents).

What distribution would the values of the first significant digit of the numbers 
take? e.g. if the numbers are all written in scientific notation: 
$$
\begin{align}
1.3&\times10^{1}\\
7.6&\times10^{-2}\\
5.1&\times10^{2}\\
 &\vdots \\
9.0&\times10^{-1}\\
2.3&\times10^{3}\\
\end{align}
$$
What will the histogram of the numbers in front of the "$.$" look like? 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE,warnings=FALSE}
curve(dnorm(round(x),5,2),from=0.5,to=9.5,lwd=2, type='s',ylim=c(0,0.4),
        xlab='First Significant Digit', ylab='PDF',col='blue3')
curve(dunif(round(x),1,9),from=0.5,to=9.5,lwd=2,add=TRUE,type='s',col='red3')
curve(dexp(round(x),1/3),from=0.5,to=9.5,lwd=2,add=TRUE,type='s',col='green3')

```
<!--}}}--> 

# The Pareto Distribution <!--{{{--> 

The last distribution that we explore in this section is the Pareto distribution, 
which is a probability function that follows a truncated power-law. The Pareto 
probability distribution is defined as: 
$$
p(x) = 
\begin{cases} 
\frac{\alpha x^\alpha_m}{x^{\alpha+1}} & x\ge x_m \\
0 & x<x_m
\end{cases}
$$
for $x_m>0$. 

The Pareto distribution is named for Vilfredo Pareto, who first used the 
probability distribution to demonstrate wealth disparity. We can repeat this 
demonstration quite simply. 

Consider $1,000$ people who each have $\$1,000$. We take two of these people at 
random, and give a fixed fraction of their total worth to the other (think of 
this a transaction of some kind). We then repeat this process $100,000$ times. 

```{r,  fig.height=4, fig.width=6, out.width='90%',eval=TRUE,echo=FALSE,warnings=FALSE}
worth<-rep(1E3,1E3)
for(i in 1:1e5){
  to.switch<-sample(1E3,2)
  transaction<-runif(1)*worth[to.switch[1]]
  worth[to.switch[1]]<-worth[to.switch[1]]-transaction
  worth[to.switch[2]]<-worth[to.switch[2]]+transaction
}
magplot(density(worth, kern='rect', bw=100/sqrt(12),n=1e3),lwd=2, 
        xlab='Net Worth per person', ylab='PDF',col='blue3', majorn=3)
abline(v=1e3,col='red3',lty=3)
```

Power-law functions are reasonably common in natural sciences, as many natural processes 
tend toward the power law. The represent relationships between parameters where 
a relative change in one quantity produces a proportional relative change in 
another quantity, regardless of the magnitude of the various parameters. 
Notable astronomical examples include Keplers third law, and the stellar initial 
mass function. Notable physical examples include the Stefan-Boltzman law and 
all inverse-square laws (electrostatics, gravitation, etc). 

One interesting demonstration of the power law is in Benfords Law. This is not 
directly related to physics or astronomy (unless you are doing something _very_
questionable with your grant applications...).

Benfords Law, also called the first-digit law, states that the leading digit of 
listings of numbers that span multiple decades (i.e. many orders of magnitude) 
will tend to be preferentially valued $1$; and the probability of finding larger 
values is inversely proportional to the digits value: 
$$
P(D)=\log_{10}\left(1+\frac{1}{D}\right).
$$

```{r}
x<-10^runif(1e4,-3,3)
y<-gsub('0','',gsub(fixed=T,'.','',as.character(x)))
y<-helpRfuncs::vecsplit(y,"",1)
maghist(as.numeric(y),breaks=1:10-0.5,verbose=F,col='blue3',freq=F,
xlab='Leading digit',ylab='PDF')
lines(col='red',lty=2,1:9,log10(1+1/(1:9)),lwd=2)
legend('topright',inset=0.1,legend=c('1000 values in 10^[-3,3]','Benfords Law'),
       pch=c(15,NA),lty=c(NA,2),col=c('blue3','red'))
```



The Pareto Distribution has properties: 

> + Expectation 

$$\mathbb{E}[X]=
\begin{cases}
\infty & \alpha \leq 1 \\
\frac{\alpha x_m}{\alpha-1} & \alpha > 1 \\
\end{cases}
$$

> + Variance 

$$
\textrm{var}[X]=
\begin{cases}
\infty & \alpha \in (1,2] \\
\left(\frac{x_m}{\alpha-1}\right)^2 \frac{\alpha}{\alpha-2} & \alpha > 2 \\
\end{cases}
$$


<!--}}}--->
